Statically Typed String Sanitation Inside a Python

### 2014-10-06 sanitation-psp14/sanitation-psp14-short.tex
typo
Updates to Sec2 of PSP14

We observe that many such sanitization techniques can be understood using \emph{regular languages} \cite{cinderella}. For example, \verb|name| must be a string in the language described by the regular expression \verb!([^"\]|(\")|(\\))*! -- a sequence of characters other than quotation marks and backslashes; these can only appear escaped. This concrete syntax for regular expression patterns can be understood to desugar, in a standard way, to the syntax for regular expressions shown in Figure \ref{fig:regex}, where $r \cdot r$ is sequencing and $r + r$ is  disjunction. We will work with this ``core'' for simplicity.

@@

\section{Regular String Types, Minimally}\label{calculus}
\noindent
%This will serve as the source language for our translation to a calculus with only standard strings and regular expressions, defined in Section~\ref{sec:tr}. 
This section is organized as follows:
\begin{itemize}
\item Section \ref{sec:rs} describes $\lambda_{RS}$. We also give proof outlines of  type safety and correctness theorems and relevant propositions about regular languages.
\item Section \ref{sec:p} describes a simple target language, $\lambdap$, with a minimal regular expression library. In Section \ref{atlang}, we will take Python to be such a language.
\item Section \ref{sec:tr} describes the translation from $\lambdas$ to $\lambdap$ and ensures the correctness result for \ref{sec:rs} is preserved under translation. 
\end{itemize}

\subsection{The Language of Regular Strings}\label{sec:rs}
In this section, we define a minimal typed lambda calculus with regular string types called $\lambda_{RS}$. The syntax of $\lambda_{RS}$ is specified in Figure \ref{fig:glambdas}, its static semantics in Figure \ref{fig:slambdas} and its evaluation semantics in Figure \ref{fig:dlambdas}.\footnote{For convenience, a single sheet containing Figures \ref{fig:glambdas}-\ref{fig:tr} is available at http://nfulton.org/strings/printout.pdf.}

There are two type constructors in $\lambdas$, $\rightarrow$ and $\textsf{stringin}$. Arrow types classify functions, which are introduced via lambda abstraction, $\lambda x.e$, and can be applied, written $e(e)$, in the usual way \cite{pfpl}. Regular string types are of the form $\stringin{r}$, where $r$ is a regular expression. Values of such regular string types take the form $\strin{s}$, where $s$ is a string (i.e. $s \in \Sigma^\star$, defined in the usual way). The rule \textsc{S-T-Stringin-I} statically guarantees that $s \in \lang{r}$.
 %The remaining operations on terms of type $\stringin{r}$ preserve this property. %The remaining operators operate on regular strings.

@@

The \textsc{S-T-Concat} rule is the simplest example of our approach. The rule is sound because the result of concatenating two strings will always be in the sequential composition two regular expressions matching the respective strings. The rule therefore relates string concatenation to sequential composition of regular expressions.

Whereas concatenation allows the construction of large strings from smaller strings, \textsc{S-T-Case} allows the decomposition, or elimination, of large strings into smaller strings. Intuitively, this rule ``peels off" the first character, and then performs a specified operation on the first character and the remaining string.

In terms of strings, \textsc{S-T-Case} is essentially the ubiquitious substring operation.

The definition of \textsc{S-T-Case} is subtle because even simple string operations sometimes require complicated operations on regular expressions. Peeling off the first character of a string is deterministic -- the first character is always one, concrete value. However, a regular expression describing the first character of any matching string might recognize multiple concrete values. For instance, the first character of ``CMU" is always ``C"; however, if the expression associated with this string is CMU$+$KIT, then the expression describing the first character is C$+$K.
Therefore, an operation on languages corresponding to case analysis on strings must consider one-character prefixes (a head) and their corresponding suffixes (a tail).

Fortunately, regular expression derivatives \cite{bowzer} conveniently capture this intuition.
The regular expression recognizing any one-character prefix of the strings in $\lang{r}$ is easily defined.
The one-character prefix definition combined with derivatives gives a concise definition of the
remaining string.

@@

Given this definition of $\lhead{r}$, regular expression derivatives provide a useful tool for defining $\ltail{r}$.
\begin{defn}[Brzozowski's Derivative]\label{def:derivative}
  The \emph{derivative of $r$ with respect to $s$} is denoted by $\delta_s(r)$
  and is $\delta_s(r) = \{t | st \in \lang{r}\}$.
\end{defn}

@@

The \textsc{S-T-Case-Concat} rule, which is defined in terms of these operations, relates the result of ``peeling off" the first character of a string to regular expression derivatives.

@@

The $\lambdas$ language supports two forms of coercion. Safe coercion only allows coercion between 
strings types which are guaranteed to be safe. This form of coercion is useful because the replacement operation (introduced below in Section~\ref{sec:replace}) is often more conservative than absolutely necessary.
Conversely, unsafe coercion -- which we refer to as a checked coercion -- 
allows for potential unsafe type casts. Our semantics for check ensures that only safe values are used. Whenever a value is determined to be unsafe at runtime, a default value $e_2$ is used instead.

Summarily, 
 \textsc{S-T-SafeCoerce} allows only safe coercions between string types by
    expoiting the decidability of language inclusion, while
\textsc{S-T-Check} allows casts between strings that cannot be guaranteed at compile time, but inserts a runtime check.

An alternative to coercion is subtyping for constrained string types, which applies coercion implicitly. 
In practice, subtyping is crucial to the usability of our system due to the nature of the replacement operator.
For simplicity, we do not include this relation; however, subtyping for constrained strings is present in \cite{fulton12,fulton13} and that treatment extends to this system. The implementation discussed in Section \ref{atlang} provides subtyping between constrained string types.

\subsubsection{Replacement}\label{sec:replace}

The premier operation for manipulating strings in $\lambdas$ is string substitution.
On strings, the string substitution operator is familiar, and roughly analogous to \lstinline{str_replace} in PHP and \lstinline{String.replace} in Java.
The definition of \textsc{S-T-Replace} requires a definition of replacement for regular expressions (or languages) corresponding to string replacement.
In this section, we define this operation.

Both string and language replacement are defined extra-linguistically.
The system $\lambdas$ is defined in terms of these functions.
The function $\lsubst{r}{s_1}{s_2}$ replaces every substring of $s_1$ matching $r$ with $s_2$.
In an analogous manner, the function $\lreplace{r}{r_1}{r_2}$ replaces every sublanguage of $r_1$ matching $r$ with $r_2$.
If the intuition for this operation is not clear, it may be helpful to think in terms of replacing sub-automata.

Throughout this section, we fix an alphabet $\Sigma$ over which strings $s$ and
regular expressions $r$ are defined. We use $\lang{r}$ to refer to the
language recognized by the regular expression $r$.

@@

\begin{defn}[$\tt{subst}$]
The relation $\lsubst{r}{s_1}{s_2} = s$ produces a string $s$ in which all substrings of $s_1$ matching $r$ are replaced with $s_2$.
\end{defn}

A proper definition of ${\sf lreplace}$ would give an rewrite system with correctness and termination proofs, which is beyond the scope of this paper
on security-motivated type system extensions. Instead, we provide an abstract definition of the relation and state necessary propositions.

\begin{prop}[Closure] \label{thm:total}
  If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular expressions, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
\end{prop}
\begin{proof}
Algorithms for the inclusion problem may be adopted to identify any sublanguage $x \subseteq r$ of $r_1$.
The language $x_{pre}$ can be computed by taking total derivatives until the remaining language
equals $x$; $x_{post}$ can be computed in a similar way after reversal.
Then $r'$ is $x_{pre}r_2x_{post}$.
\end{proof}

  $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{s_1}{s_2}}$.
\end{prop}
\begin{proof}
The proposition follows from the definitions of subst and lreplace; note that language substitutions over-approximate string substitutions.
\end{proof}

>>>>

We observe that many such sanitization techniques can be understood using \emph{regular languages} \cite{cinderella}. For example, \verb|name| must be a string in the language described by the regular expression \verb!([^"\]|(\")|(\\))*! -- a sequence of characters other than quotation marks and backslashes; these can only appear escaped. This syntax for regular expression patterns can be understood to desugar, in a standard way, to the syntax for regular expressions shown in Figure \ref{fig:regex}, where $r \cdot r$ is sequencing and $r + r$ is  disjunction. We will work with this ``core'' for simplicity.

@@

\section{Regular String Types, Minimally}\label{calculus}
\noindent
%This will serve as the source language for our translation to a calculus with only standard strings and regular expressions, defined in Section~\ref{sec:tr}. 
This section is organized as follows:
\begin{itemize}
\item Section \ref{sec:rs} describes $\lambda_{RS}$. We also give proof outlines of  type safety and correctness theorems and relevant propositions about regular languages.
\item Section \ref{sec:p} describes a simple target language, $\lambdap$, with a minimal regular expression library. In Section \ref{atlang}, we will take Python to be such a language.
\item Section \ref{sec:tr} describes the translation from $\lambdas$ to $\lambdap$ and ensures the correctness result for \ref{sec:rs} is preserved under this translation. 
\end{itemize}

\subsection{The Language of Regular Strings}\label{sec:rs}
In this section, we define a minimal typed lambda calculus with regular string types called $\lambda_{RS}$. The syntax of $\lambda_{RS}$ is specified in Figure \ref{fig:glambdas}, its static semantics in Figure \ref{fig:slambdas} and its evaluation semantics in Figure \ref{fig:dlambdas}.\footnote{For convenience, a single sheet containing Figures \ref{fig:glambdas}-\ref{fig:tr} is available in the accompanying technical report \todo{TR}.}

There are two type constructors in $\lambdas$, $\rightarrow$ and $\textsf{stringin}$. Arrow types classify functions, which are introduced via lambda abstraction, $\lambda x.e$, and can be applied, written $e(e)$, in the usual way \cite{pfpl}. Regular string types are of the form $\stringin{r}$, where $r$ is a regular expression. Values of such regular string types take the form $\strin{s}$, where $s$ is a string (i.e. $s \in \Sigma^\star$, defined in the usual way). The rule \textsc{S-T-Stringin-I} requires that $s \in \lang{r}$.
 %The remaining operations on terms of type $\stringin{r}$ preserve this property. %The remaining operators operate on regular strings.

@@

The \textsc{S-T-Concat} rule is the simplest example of our approach. The rule is sound because the result of concatenating two strings, the first in $\lang{r_1}$ and the second in $\lang{r_2}$, will always be in the language $\lang{r_1\cdot r_2}$. The rule therefore relates string concatenation to sequential composition of regular expressions.

Whereas concatenation allows the construction of large strings from smaller strings,  $\strcase{e}{e_0}{x,y.e_1}$  allows the decomposition, or elimination, of large strings into smaller strings. Intuitively, this operation branches based on whether a string is empty or not, exactly analagous to case analysis on lists in a functional language. The branch for a non-empty string ``peels off" the first character, binding it and the remainder of the string to specified variables. The evaluation rules \textsc{S-E-Case-$\epsilon$} and \textsc{S-E-Case-Concat} express this semantics. This construct can be used to implement a standard string indexing operation as a function, given a suitable definition of natural numbers (omitted).

The regular expressions describing the first character and remaining characters of a matching string in the language $\lang{r}$ must be determined to correctly specify the static semantics. For instance, the first character of a string in the language $\lang{(A\cdot B\cdot C)+(D\cdot E\cdot F)}$ is in the language $\lang{A+D}$ and the remaining characters are in the language $\lang{(B\cdot C)+(E \cdot F)}$.

The regular expression recognizing any one-character prefix of the strings in $\lang{r}$ is easily defined.

@@

Given this definition of $\lhead{r}$, regular expression derivatives \cite{bowzer} provide a useful tool for defining $\ltail{r}$.
\begin{defn}[Brzozowski's Derivative]\label{def:derivative}
  The \emph{derivative of $r$ with respect to $s$} is denoted by $\delta_s(r)$
  and is $\delta_s(r) = \{t | st \in \lang{r}\}$.
\end{defn}

@@

The \textsc{S-T-Case} rule, which is defined in terms of these operations, thus relates the result of ``peeling off" the first character of a string to regular expression derivatives.

@@

The $\lambdas$ language supports two forms of coercion. Safe coercions, written $\rcoerce{r}{e}$, allow passage to a ``smaller'' regular language. Such coercions will always succeed. 
% This form of coercion is useful because the replacement operation (introduced below in Section~\ref{sec:replace}) is often more conservative than absolutely necessary.
Conversely, checked coercions, written $\rcheck{r}{e_0}{x.e_1}{e_2}$, 
allow for passage to regular languages that are not necessarily smaller. Checked coercions branch based on whether the coercion succeeded or not.

The rule 
 \textsc{S-T-SafeCoerce} checks for language inclusion, which we write $\lang{r_1} \subseteq \lang{r_2}$. Language inclusion is a decidable relation. As a result, the rule \textsc{S-E-SafeCoerce} does not need to perform any checks. The rule 
\textsc{S-T-Check} does not perform any static checks on the two languages involved, only checking that the two branches have the same type. The checks are performed dynamically, shown in the rules \textsc{S-E-Check-Ok} and \textsc{S-E-Check-NotOk}.

In our calculus, both forms of coercion are explicit. For safe coercions, it is often useful for the coercion to be performed implicitly. This can be seen as a form of  subtyping for regular string types \todo{cite?}. 
In practice, subtyping is crucial to the usability of our system due to the nature of the replacement operator.
For simplicity, we do not include this relation; however, subtyping for regular strings is present in previous treatments of our work \cite{fulton12,fulton13}. The implementation discussed in Section \ref{atlang} also provides subtyping between regular string types.

\subsubsection{Replacement}\label{sec:replace}
The premier operation for working with regular strings in $\lambdas$ is  replacement, written $\sisubst{r}{e_1}{e_2}$. 
It behaves analogously to \lstinline{str_replace} in PHP and \lstinline{String.replace} in Java, differing in that the replacement pattern $r$ must be statically given. The evaluation rule \textsc{S-E-replace} is defined via the relation $\lsubst{r}{s_1}{s_2}=s$. For simplicity, we do not inductively define this standard relation.

\begin{defn}[$\tt{subst}$]
The relation $\lsubst{r}{s_1}{s_2} = s$ produces a string $s$ in which all substrings of $s_1$ in $\lang{r}$ are replaced with $s_2$.
\end{defn}

The typing rule \textsc{S-T-Replace} requires computing the regular language of the string resulting from this operation given the languages of $s_1$ and $s_2$, written as the relation $\lreplace{r}{r_1}{r_2}={r'}$. 
Given an automata-oriented interpretation of regular languages, it may be helpful to think in terms of replacing sub-automata.

A complete definition of ${\sf lreplace}$ would also give an rewrite system with correctness and termination proofs, which is beyond the scope of this paper. Instead, we provide an abstract definition of the relation and state necessary properties.

\begin{prop}[Closure] \label{thm:total}
  If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular languages, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
\end{prop}
\begin{proof}
Algorithms for the inclusion problem may be adopted to identify any sublanguage $x \subseteq r$ of $r_1$.
The language $x_{pre}$ can be computed by taking total derivatives until the remaining language
equals $x$; $x_{post}$ can be computed in a similar way after reversal.
Then $r'$ is $x_{pre}r_2x_{post}$.
\end{proof}

  $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
\end{prop}
\begin{proof}
The proposition follows from the definitions of subst and lreplace; note that language substitutions over-approximate string substitutions.
\end{proof}

@@
